\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

title: "Final Project"
author: "Fiona Price"
date: "2024-11-11"
output: html_document
editor_options: 
  chunk_output_type: console
  
---

```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(here)
library(tidyverse)
library(readxl)
library(sf)
library(mapview)
```

```{r}
#Check file path
here()
#Load in data. Only looking at the fourth sheet.
energy_data <- read_excel(here("./Data/Raw/egrid2022_data.xlsx"), sheet = 4)
#Our table has sub-headers, which we don't need, so we can get rid of that row.
energy_data <- energy_data[-1, ]

#Load in spatial data
nc_counties <- st_read(
  here("./Data/Spatial/North_Carolina_State_and_County_Boundary_Polygons.shp")) 
#Reveal the CRS
st_crs(nc_counties)$epsg
```

```{r}
nc_energy <- energy_data %>% 
  filter(energy_data$`Plant state abbreviation` == "NC")

nc_plants<- nc_energy %>% 
    st_as_sf(coords = c("Plant longitude","Plant latitude"), 
             crs=4269) 

nc_plants$`Plant nameplate capacity (MW)` <- as.numeric(nc_plants$`Plant nameplate capacity (MW)`)
class(nc_plants$`Plant nameplate capacity (MW)` )

#Color by fuel category
ggplot() +
  geom_sf(data = nc_counties, fill = "grey") +
  geom_sf(data = nc_plants, aes(color = nc_plants$`Plant primary fuel category`)) +
  labs(color = "Fuel Category")
#Mapview option
mapview(nc_plants, zcol = "Plant primary fuel category") +
  mapview(nc_counties, col.regions = "grey")


#Size by capacity
#Ggplot option
ggplot() +
  geom_sf(data = nc_counties, fill = "grey") +
  geom_sf(data = nc_plants, aes(size = nc_plants$`Plant nameplate capacity (MW)`)) + 
  theme(legend.position = "none")
#Mapview option
mapview(nc_plants, cex = "Plant nameplate capacity (MW)",
        zcol = "Plant nameplate capacity (MW)") +
  mapview(nc_counties, col.regions = "grey")

```{r setup, include=FALSE}
# Set your working directory

# Load your packages

# Set your ggplot theme

# Load your datasets

author: "Asreeta Ushasri"


#view(energy_data)
#view(nc_energy)
#view(nc_counties)
#view(nc_plants)

nc_energy_numeric <- nc_energy %>%
  mutate(`Plant_Cap` = as.numeric(`Plant nameplate capacity (MW)`)) %>% 
  drop_na()


```

```


# Rationale and Research Questions



\newpage

# Dataset Information



\newpage

# Exploratory Analysis 



\newpage

# Analysis



## Question 1: <insert specific question here and add additional subsections for additional questions below, if needed>

## Question 2: 




\newpage

# Summary and Conclusions


\newpage

# References
<add references here if relevant, otherwise delete this section> 
