\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

title: "Examining Plant Capacity"
author: "Asreeta Ushasri"
date: "2024-11-11"
output: html_document
editor_options: 
  chunk_output_type: console
  
---

```{r setup, include=FALSE}

#Load packages

library(dplyr)
library(ggplot2)
library(here)
library(tidyverse)
library(readxl)
library(sf)
library(mapview)

#Check working directory

getwd()
here()


```

```{r}
#Load in the power plants data, code by Fiona Price

energy_data <- read_excel(here("./Data/Raw/egrid2022_data.xlsx"), sheet = 4)

#Remove subheaders

energy_data <- energy_data[-1, ]

#Load in spatial data

nc_counties <- st_read(
  here("./Data/Spatial/North_Carolina_State_and_County_Boundary_Polygons.shp")) 
#Reveal the CRS
st_crs(nc_counties)$epsg

```

```{r}
# Set up NC datasets
# Code written by Fiona Price

nc_energy <- energy_data %>% 
  filter(energy_data$`Plant state abbreviation` == "NC")

nc_plants<- nc_energy %>% 
    st_as_sf(coords = c("Plant longitude","Plant latitude"), 
             crs=4269) 

nc_plants$`Plant nameplate capacity (MW)` <- as.numeric(nc_plants$`Plant nameplate capacity (MW)`)

class(nc_plants$`Plant nameplate capacity (MW)` )

#author: "Asreeta Ushasri"

default_theme <- theme_light() +
  theme(panel.background = element_rect(color = "white"),
  legend.key = element_rect(color = "white"),
    legend.position = 'right',
    complete = TRUE)

theme_set(default_theme)

view(energy_data)
view(nc_energy)
view(nc_counties)
view(nc_plants)

nc_energy_total_cap <- nc_energy %>%
  group_by(`Plant primary fuel category`) %>%
  summarise(total_capacity = sum(as.numeric(`Plant nameplate capacity (MW)`)))

view(nc_energy_total_cap)

ggplot(data = nc_energy_total_cap) +
  geom_bar(aes(x = `Plant primary fuel category`,
               y = `total_capacity`), fill = 'darkblue',
           stat = "identity") +
  labs(title = "Power Plant Capacity in North Carolina",
       subtitle = "Asreeta Ushasri",
       x = "Plant Primary Fuel Category",
       y = "Nameplate Capacity (MW)")

```

```


author: "Asreeta Ushasri"



\newpage

# Dataset Information

The Annual Technology Baseline data set provides information on electricity generation in the United States. The National Renewable Energy Labratory publishes this data to describe capacity and geographic distribution of different energy resources. The 2022 data is the most recent data (published in 2024), because it requires time to process and handle previous data in a workbook.

This project primary utilizes the Plant 2022 and the Generator 2022 data sheets. The Plant 2022 data sheet describes the capacity, location, and primary fuel source in each power plant. The Generator 2022 worksheet focuses on individual generators. Power plants often have multiple generators of various fuel types, (such as natural gas combined cycle plants), leading to two separate data sheets to describe this informaiton.




\newpage

# Exploratory Analysis 



\newpage

# Analysis



## Question 1: <insert specific question here and add additional subsections for additional questions below, if needed>

## Question 2: 




\newpage

# Summary and Conclusions


\newpage

# References
<add references here if relevant, otherwise delete this section> 
