---
title: "Data Wrangling/Processing"
author: "Fiona Price, Asreeta Ushasri, Summer Heschong"
date: "2024-11-30"
output: html_document
editor_options: 
  chunk_output_type: console
---

\tableofcontents


```{r setup, include=FALSE}
library(dplyr)
library(ggplot2)
library(here)
library(tidyverse)
library(readxl)
library(sf)

```

#1. Energy Data

##A. Load in Raw EGRID 2022 Data

```{r }
#Check file path
here()
#Load in data. Only looking at the fourth sheet.
energy_data <- read_excel(here("./Data/Raw/egrid2022_data.xlsx"), sheet = 4)
#Our table has sub-headers, which we don't need, so we can get rid of that row.
energy_data <- energy_data[-1, ]

```

##B. Select Columns Related to State, Location, Primary Fuel Category, Capacity, and Emissions

```{r}
#Find list of all columns
colnames(energy_data)
#Select only the relevant columns
energy_data_selected <- energy_data %>% 
  select('Data Year', 'Plant state abbreviation', 'Plant name', 'Utility name', 
         'Plant county name', 'Plant latitude', 'Plant longitude', 'Number of units',
         'Number of generators', 'Plant primary fuel category', 
         'Plant nameplate capacity (MW)', 'Plant annual net generation (MWh)',
         'Plant annual NOx emissions (tons)', 'Plant annual SO2 emissions (tons)',
         'Plant annual CO2 emissions (tons)', 'Plant annual CH4 emissions (lbs)',
         'Plant annual N2O emissions (lbs)', 'Plant annual CO2 equivalent emissions (tons)')
```

##C. Filter for North Carolina, the focus of this project

```{r}
#Select for only NC
NC_energy <- energy_data_selected %>% 
  filter(energy_data_selected$`Plant state abbreviation` == "NC")
```

##D. Adjusting Date Columns

```{r}
#Check class of year column
class(NC_energy$`Data Year`)
#Currently a character. Change to date.
NC_energy$`Data Year` <-  as.Date(NC_energy$`Data Year`, format = "%Y")

```

#2. Environmental Justice Community Data
##A. Uploading Raw Environmental Justice Data, summarizing percentages

```{r} 
#Load in CEJST data
cejst_data <- read.csv(here("./Data/Raw/CEJST_data.csv"))
#Filter for NC
NC_cejst <- cejst_data %>% 
  filter(cejst_data$State.Territory == "North Carolina")
#Remove the word "County" from the county column
NC_cejst$County.Name <- gsub("\\s*[Cc]ounty", "", NC_cejst$County.Name)
#Find a total county average for relevant columns
counties_summarized <- NC_cejst %>% 
  group_by(County.Name) %>% 
  summarise(percent_black = mean(Percent.Black.or.African.American.alone),
            percent_na = mean(Percent.American.Indian...Alaska.Native),
            percent_asian = mean(Percent.Asian),
            percent_hawaiian = mean(Percent.Native.Hawaiian.or.Pacific),
            percent_two = mean(Percent.two.or.more.races),
            percent_white = mean(Percent.White),
            percent_hispanic = mean(Percent.Hispanic.or.Latino),
            disadvantaged = mean(Percentage.of.tract.that.is.disadvantaged.by.area),
            population = sum(Total.population),
            percent_poverty = mean(Percent.of.individuals.below.200..Federal.Poverty.Line..percentile.),
            PM2.5 = mean(PM2.5.in.the.air))

```

##B. Joining County Data into Environmental Justice Data

```{r}
#Rename the county column name in the energy df to match the CEJST df
NC_energy <- NC_energy %>% 
  rename("County.Name" = "Plant county name")
#Join CEJST and energy data by county name
NC_joined <- merge(NC_energy, counties_summarized, by = "County.Name", all = FALSE)
```

##3. Saving Processed Environmental Justice Data

```{r}
#Save joined data as a file in the processed data folder.
write.csv(NC_joined, row.names = FALSE, 
         file = "./Data/Processed/NC_joined.csv")
```

##4. Saving Processed Energy Capacity Data

```{r}
nc_energy_total_cap <- NC_energy %>%
  group_by(`Plant primary fuel category`) %>%
  summarise(total_capacity = sum(as.numeric(`Plant nameplate capacity (MW)`)))

write.csv(nc_energy_total_cap, row.names = FALSE, 
         file = "./Data/Processed/NC_Total_Capacity.csv")
```

##5. Saving Processed Energy Emissions Data

```{r}
nc_energy_emissions <- NC_energy %>%
  group_by(`Plant primary fuel category`) %>% summarise(total_CO2e = sum(as.numeric(`Plant annual CO2 equivalent emissions (tons)`)))

write.csv(nc_energy_emissions, row.names = FALSE, 
         file = "./Data/Processed/NC_Energy_Emissions.csv")
```

##6. Saving Processed North Carolina Power Plant Data

```{r}

write.csv(NC_energy, row.names = FALSE, 
         file = "./Data/Processed/NC_PowerPlants.csv")

```


```

