---
title: "Electricity Generation and Environmental Justice Communities in North Carolina"
author: "Fiona Price, Summer Heschong, and Asreeta Ushasri"
date: "2024-12-12"
output: html_document
editor_options: 
  chunk_output_type: console
---

\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

```{r setup, include=FALSE}
# Load your packages

library(dplyr)
library(ggplot2)
library(here)
library(tidyverse)
library(readxl)
library(sf)
library(mapview)
library(viridis)
library(readr)
library(tigris)


# Check your working directory

getwd()
here()

# Set your ggplot theme

default_theme <- theme_light() +
  theme(panel.background = element_rect(color = "white"),
  legend.key = element_rect(color = "white"),
    legend.position = 'right',
    complete = TRUE)

theme_set(default_theme)

# Load your datasets

energy_data <-
  read_csv("./Data/Processed/NC_PowerPlants.csv")

nc_energy_total_cap <-
  read_csv("./Data/Processed/NC_Total_Capacity.csv")

nc_energy_emissions <-
  read_csv("./Data/Processed/NC_Energy_Emissions.csv")

nc_plants <- energy_data %>% 
    st_as_sf(coords = c("Plant longitude","Plant latitude"), 
             crs=4269) 

nc_plant_emissions <- energy_data %>% 
    st_as_sf(coords = c("Plant longitude","Plant latitude"), 
             crs=4269)

nc_EJ_powerplants <- read_csv("./Data/Processed/NC_joined.csv")

EJ_NC_sf <- st_read("./Data/Processed/NC_Env_Justice.gpkg") 

#County spatial data

nc_counties <- st_read(
  here("./Data/Spatial/North_Carolina_State_and_County_Boundary_Polygons.shp")) 

#Check the CRS
st_crs(nc_counties)$epsg

```


# Rationale and Research Questions

1. What is the distribution and capacity of power plants generating electricity in North Carolina?
2. What are the Environmental Justice implications of electricity power generation in North Carolina? Are power plants primarily located in low-income and disadvantaged communities?
3. Do Environmental Justice communities face impact from carbon dioxide equivalent emissions due to power plants?

\newpage

# Dataset Information

The eGRID 2022 data set provides information on electricity generation in the United States. The U.S. EPA publishes this data to describe capacity and geographic distribution of different energy resources. The 2022 data is the most recent data (published in 2024), because it requires time to process and compile this national data into a workbook.

This project utilizes the Plant 2022 and the Generator 2022 data sheets. The Plant 2022 data sheet describes the capacity, location, and primary fuel source of every power plant in the United States. The Generator 2022 worksheet focuses on individual generators. Power plants often have multiple generators of various fuel types, (such as natural gas combined cycle plants). For this reason, the U.S. EPA includes a separate worksheet for generators specifically.

Lastly, this project incorporates spatial data to understand the geographic relation between plants, as well as between plants and EJ communities. The spatial data is from NC OneMap; it includes a vector polygon dataset delineating official boundaries of the North Carolina counties (in addition to the North Carolina state boundary as well). 

\newpage

# Exploratory Analysis 
```{r Plant Primary Fuel Cateory and Capacity Plots}
#view(nc_energy_total_cap)
#colnames(energy_data)

NC_Plant_Cap_Plot <- ggplot(data = nc_energy_total_cap) +
  geom_bar(aes(x = `Plant primary fuel category`,
               y = `total_capacity`), fill = 'darkblue',
           stat = "identity") +
  labs(title = "Power Plant Capacity in North Carolina",
       x = "Plant Primary Fuel Category",
       y = "Nameplate Capacity (MW)")

print(NC_Plant_Cap_Plot)

#Find the frequency of each type of fuel category
plant_counts <- nc_plants %>%
  group_by(`Plant primary fuel category`) %>%
  summarize(Count = n())
#Plot it
FuelType_Count_Plot <- 
  ggplot(plant_counts, aes(x = `Plant primary fuel category`, y = Count, 
                         fill = `Plant primary fuel category`)) +
  geom_bar(stat = "identity") +  
  geom_text(aes(label = Count), vjust = -0.5, size = 2.5) +
  labs(title = "Number of Plants by Fuel Type", x = "Fuel Type", y = "Number of Plants") +
  theme(axis.text.x = element_text(angle = 45, hjust = 0.3, size = 8)) +
  scale_fill_viridis(discrete = TRUE, option = "C")

print(FuelType_Count_Plot)

```


```{r Spatial Analysis}
#Mapping capacity of electricity generation in NC

#Convert capacity to numeric
nc_plants$`Plant nameplate capacity (MW)` <-
  as.numeric(nc_plants$`Plant nameplate capacity (MW)`)
#Check class
class(nc_plants$`Plant nameplate capacity (MW)` )

#Color by fuel category
Plant_Location_Color_Plot <- 
  ggplot() +
  geom_sf(data = nc_counties, fill = "grey") +
  geom_sf(data = nc_plants, aes(color = nc_plants$`Plant primary fuel category`)) +
  labs(color = "Fuel Category",
       title = "Spatial Distribution of Power Plants in NC") +
  theme(legend.text = element_text(size = 8),
        legend.title = element_text(size = 8)) +
   scale_color_viridis_d(option = "C")

print(Plant_Location_Color_Plot)

#Mapview option - allows for interactive use
Plant_Location_Color_Map <- 
  mapview(nc_plants, layer.name = "NC Power Plants", 
          zcol = "Plant primary fuel category",
          col.regions = viridis::viridis(length(unique(nc_plants$`Plant primary fuel category`)), 
                                         option = "C")) +
  mapview(nc_counties, layer.name = "NC Counties", col.regions = "grey") 

print(Plant_Location_Color_Map)

#Size by capacity
#Ggplot option
Plant_Location_Size_Plot <- 
  ggplot() +
  geom_sf(data = nc_counties, fill = "grey") +
  geom_sf(data = nc_plants, aes(size = nc_plants$`Plant nameplate capacity (MW)`,
                                color = "red")) +
  labs(title = "Spatial Distribution of Power Plants in NC, Sized by Capacity",
       size = "Capacity (MW)") +
  default_theme +
  theme(legend.position = "bottom")

print(Plant_Location_Size_Plot)

#Mapview option
Plant_Location_Size_Map <- 
  mapview(nc_plants, cex = "Plant nameplate capacity (MW)", 
          zcol = "Plant primary fuel category",
          col.regions = viridis::viridis(length(unique(nc_plants$`Plant primary fuel category`)), 
                                         option = "C"),
          layer.name = 'Plant Primary Fuel Category') +
  mapview(nc_counties, col.regions = "grey", layer.name = 'Counties')

print(Plant_Location_Size_Map)
```

```{r EJ Analusos}
#map which counties are classified as disadvantaged

mapview(EJ.NC_sf, zcol = 'Identified.as.disadvantaged',
        layer.name = 'County Disadvantaged Status')

```


\newpage

# Analysis

## Question 1: What is the distribution and capacity of power plants generating electricity in North Carolina?
###Question 1a: Which type of plant has the highest capacity?
The exploratory analysis above investigates which plant fuel category has the highest capacity. In total, gas, solar, nuclear, and coal have the highest capacities. However, this does not reflect the average capcity for each fuel type, as shown in figure X below. As displayed in this graph, nuclear has the highest average capacity, followed by coal, gas, wind, hydro, other, biomass, oil, and solar. Although has has the highest total capacity, each individual plant has a lower average capacity than nuclear and coal. This suggests there is a greater quantity of gas plants than nuclear or coal.

###Question 1b: Which type of plant is the most common?
Graph X shows the number of each type of plant. Solar is by far the most common, followed by oil, hydro, biomass, and gas. The reason solar does not have the highest total capacity in MW for NC (despite being the most common) is that it has the lowest average capacity for each plant (Graph X).

###Question 1c: Where are most plants located?
The ten counties with the most power plants are, in order, Robseon, Duplin, Wayne, Nash, Cleveland and Johnston (tied at 29), and Bladen, Catawba, Halifax, and Randolph (all tied at 17). To understand the breakdown of power plants by fuel type in the county with the most plants (Robeson), this county was isolated for analysis. The primary power plant type in Robeson County was solar (41 plants, versus biomass (2) and oil (1)).


```{r}
#Which plant fuel categories have the highest capacity?
print(NC_Plant_Cap_Plot)
#From this plot, it seems gas, solar, nuclear, and coal have the highest capacities.
#Find the average capacity of each fuel type.
average_capacity <- nc_plants %>% 
  group_by(`Plant primary fuel category`) %>% 
  summarize(avg_capacity = mean(`Plant nameplate capacity (MW)`))
#This shows that nuclear has the highest average capacity, followed by coal, gas,
#wind, hydro, other, biomass, oil, and solar. 

#Which plant fuel categories are the most common?
print(FuelType_Count_Plot)
#This plot shows that solar is by far the most common, followed by oil, hydro, 
#biomass, and gas.

#Where are most power plants in NC located?
print(Plant_Location_Color_Map)
#It is difficult to tell from this map.
county_counts <- nc_plants %>% 
  group_by(County.Name) %>% 
  summarize(plant.counts = n())
#Plot the counties with the top 10 highest number of plants.
most_plants_counties <- county_counts %>% 
  filter(plant.counts >= 17)
Plants_by_County_Plot <- ggplot(most_plants_counties, 
                aes(x = fct_reorder(`County.Name`, plant.counts, .desc = TRUE), 
                y = plant.counts, 
                fill = plant.counts)) +
  geom_bar(stat = "identity") +  
  geom_text(aes(label = plant.counts), vjust = -0.5, size = 2.5) +
  labs(title = "Number of Plants per County", x = "County", y = "Number of Plants",
       fill = "County Name") +
  default_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 0.3, size = 8)) +
  scale_fill_viridis(option = "C") 
  
print(Plants_by_County_Plot)
#This plot shows the top 10 counties with the most plants. Robeson county has
#the highest number of plants.

#Plot the breakdown of plant type for Robeson only. 
#First, filter for Robeson.
Robeson <- nc_plants %>% 
  filter(nc_plants$County.Name == "Robeson")
#Find the frequency of each type of fuel category
Robeson_fuel <- Robeson %>%
  group_by(`Plant primary fuel category`) %>%
  summarize(Count = n())

#Now plot by fuel type.
Robeson_Plot <- 
  ggplot(Robeson_fuel, aes(x = `Plant primary fuel category`, y = Count, 
                         fill = `Plant primary fuel category`)) +
  geom_bar(stat = "identity") +  
  geom_text(aes(label = Count), vjust = -0.5, size = 2.5) +
  labs(title = "Number of Plants by Fuel Type", 
       subtitle = "Robeson County",
       x = "Fuel Type", y = "Number of Plants",
       fill = "Plant Primary Fuel Category") +
  default_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 0.3, size = 8)) +
  scale_fill_viridis_d(option = "C") 

print(Robeson_Plot)
#Robeson County has primarily solar plants.
```

## Question 2: What are the Environmental Justice implications of electricity power generation in North Carolina? Are power plants primarily located in low-income and disadvantaged communities?






```{r}
#Map plants and disadvantaged counties together
mapview(EJ.NC_sf, zcol = 'Identified.as.disadvantaged',
        layer.name = 'County Disadvantaged Status') +
  mapview(nc_plants, cex = 2,
          layer.name = 'Power Plant Locations')

#view(nc_plants)

#Map plants by fuel type and disadvantaged counties
mapview(EJ.NC_sf, zcol = 'Identified.as.disadvantaged',
        layer.name = 'County Disadvantaged Status') +
  mapview(nc_plants, cex = 4, zcol = "Plant.primary.fuel.category", layer.name = 'Plant Primary Fuel Category') 

```


```{r are plant locations significant?}
#run a Chi-square test
observed <- table(EJ.NC_plus_plants$Identified.as.disadvantaged)
expected <- table(EJ.NC_sf$Identified.as.disadvantaged)

chisq_test <- chisq.test(observed, p = expected / sum(expected))
chisq_test

#filter nc_plants data set for the type of energy that is the worst for human health 
nc_unhealthy_plants <- EJ.NC_plus_plants %>%
  filter(Plant.primary.fuel.category== 'OIL' | Plant.primary.fuel.category == 'GAS' | Plant.primary.fuel.category== 'COAL' | Plant.primary.fuel.category=='NUCLEAR' |Plant.primary.fuel.category=='BIOMASS')

#Perform Chi-square test for worst types of energy to live by
observed <- table(nc_unhealthy_plants$Identified.as.disadvantaged)
expected <- table(EJ.NC_sf$Identified.as.disadvantaged)

chisq_test <- chisq.test(observed, p = expected / sum(expected))
chisq_test

```


## Question 3: Do Environmental Justice communities face impact from carbon dioxide equivalent emissions due to power plants?


```{r}
#Examine emissions by fuel type in North Carolina

#view(nc_energy_emissions)

NC_Plant_CO2e_Plot <- ggplot(data = nc_energy_emissions) +
  geom_bar(aes(x = `Plant primary fuel category`,
               y = `total_CO2e`), fill = 'darkred',
           stat = "identity") +
  labs(title = "CO2 Equivalent Emissions from
       Electricity Fuel Type in North Carolina",
       x = "Plant Primary Fuel Category",
       y = "CO2 Equivalent Emissions (Tons)")

print(NC_Plant_CO2e_Plot)

#Create map on CO2 Equivalent Emissions by Fuel Type

mapview(EJ.NC_sf, zcol = 'Identified.as.disadvantaged',
        layer.name = 'County Disadvantaged Status') +
  mapview(nc_plant_emissions,
        cex = "Plant annual CO2 equivalent emissions (tons)",
        zcol = "Plant primary fuel category",
        layer.name = "CO2 Equivalent Emissions (Tons)")

```


\newpage

# Summary and Conclusions


\newpage

# References
<add references here if relevant, otherwise delete this section> 
