---
title: "Electricity Generation and Environmental Justice Communities in North Carolina"
author: "Fiona Price, Summer Heschong, and Asreeta Ushasri"
date: "2024-12-12"
output: html_document
editor_options: 
  chunk_output_type: console
---

\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

```{r setup, include=FALSE}
# Load your packages

library(dplyr)
library(ggplot2)
library(here)
library(tidyverse)
library(readxl)
library(sf)
library(mapview)
library(viridis)
library(readr)
library(tigris)

# Check your working directory

getwd()
here()

# Set your ggplot theme

default_theme <- theme_light() +
  theme(panel.background = element_rect(color = "white"),
  legend.key = element_rect(color = "white"),
    legend.position = 'right',
    complete = TRUE)

theme_set(default_theme)

# Load your datasets

energy_data <-
  read_csv("./Data/Processed/NC_PowerPlants.csv")

nc_energy_total_cap <-
  read_csv("./Data/Processed/NC_Total_Capacity.csv")

nc_energy_emissions <-
  read_csv("./Data/Processed/NC_Energy_Emissions.csv")

nc_plants <- energy_data %>% 
    st_as_sf(coords = c("Plant longitude","Plant latitude"), 
             crs=4269) 

nc_plant_emissions <- energy_data %>% 
    st_as_sf(coords = c("Plant longitude","Plant latitude"), 
             crs=4269)

nc_EJ_powerplants <- read_csv("./Data/Processed/NC_joined.csv")

nc_plants_counties <- st_read('Data/Processed/nc_plants_EJ.gpkg')

EJ_NC_sf <- st_read("./Data/Processed/NC_Env_Justice.gpkg")

#County spatial data

nc_counties <- st_read(
  here("./Data/Spatial/North_Carolina_State_and_County_Boundary_Polygons.shp")) 

#Check the CRS
st_crs(nc_counties)$epsg

```


# Rationale and Research Questions

1. What is the distribution and capacity of power plants generating electricity in North Carolina?
2. What are the Environmental Justice implications of electricity power generation in North Carolina? Are power plants primarily located in low-income and disadvantaged communities?
3. Do Environmental Justice communities face impact from carbon dioxide equivalent emissions due to power plants?

\newpage

# Dataset Information

The eGRID 2022 data set provides information on electricity generation in the United States. The U.S. EPA publishes this data to describe capacity and geographic distribution of different energy resources. The 2022 data is the most recent data (published in 2024), because it requires time to process and compile this national data into a workbook.

This project utilizes the Plant 2022 and the Generator 2022 data sheets. The Plant 2022 data sheet describes the capacity, location, and primary fuel source of every power plant in the United States. The Generator 2022 worksheet focuses on individual generators. Power plants often have multiple generators of various fuel types, (such as natural gas combined cycle plants). For this reason, the U.S. EPA includes a separate worksheet for generators specifically.

Lastly, this project incorporates spatial data to understand the geographic relation between plants, as well as between plants and EJ communities. The spatial data is from NC OneMap; it includes a vector polygon dataset delineating official boundaries of the North Carolina counties (in addition to the North Carolina state boundary as well). 

\newpage

# Exploratory Analysis 

```{r Plant Primary Fuel Cateory and Capacity Plots}
#view(nc_energy_total_cap)
#colnames(energy_data)

NC_Plant_Cap_Plot <- ggplot(data = nc_energy_total_cap) +
  geom_bar(aes(x = `Plant primary fuel category`,
               y = `total_capacity`), fill = 'darkblue',
           stat = "identity") +
  labs(title = "Power Plant Capacity in North Carolina",
       x = "Plant Primary Fuel Category",
       y = "Nameplate Capacity (MW)")

print(NC_Plant_Cap_Plot)

#Find the frequency of each type of fuel category
plant_counts <- nc_plants %>%
  group_by(`Plant primary fuel category`) %>%
  summarize(Count = n())
#Plot it
FuelType_Count_Plot <- 
  ggplot(plant_counts, aes(x = `Plant primary fuel category`, y = Count, 
                         fill = `Plant primary fuel category`)) +
  geom_bar(stat = "identity") +  
  geom_text(aes(label = Count), vjust = -0.5, size = 2.5) +
  labs(title = "Number of Plants by Fuel Type", x = "Fuel Type", y = "Number of Plants") +
  theme(axis.text.x = element_text(angle = 45, hjust = 0.3, size = 8)) +
  scale_fill_viridis(discrete = TRUE, option = "C")

print(FuelType_Count_Plot)

```


As shown in Figure 1, natural gas generates the largest capacity of electricity power in North Carolina. The sum of natural gas power plants' nameplate capacities outranks solar, nuclear and coal. This follows broader trends in the United States. After technological advances in horizontal drilling and shale fracking in the 2010s, natural gas became economically more efficient to process. Natural gas has therefore transformed into a primary fuel source for electricity generation in the United States.

Although natural gas contributes the most to electricity capacity in North Carolina, the number of solar power plants in the state is notably high, displayed in Figure 2. This indicates that North Carolina likely hosts numerous, small solar power plants across the state. Solar power plants are small in capacity, but remain the most popular renewable resource in the state.


```{r Spatial Analysis}
#Mapping capacity of electricity generation in NC

#Convert capacity to numeric
nc_plants$`Plant nameplate capacity (MW)` <-
  as.numeric(nc_plants$`Plant nameplate capacity (MW)`)
#Check class
class(nc_plants$`Plant nameplate capacity (MW)` )

#Color by fuel category
Plant_Location_Color_Plot <- 
  ggplot() +
  geom_sf(data = nc_counties, fill = "grey") +
  geom_sf(data = nc_plants, aes(color = nc_plants$`Plant primary fuel category`)) +
  labs(color = "Fuel Category",
       title = "Spatial Distribution of Power Plants in NC") +
  theme(legend.text = element_text(size = 8),
        legend.title = element_text(size = 8)) +
   scale_color_viridis_d(option = "C")

print(Plant_Location_Color_Plot)

#Mapview option - allows for interactive use
Plant_Location_Color_Map <- 
  mapview(nc_plants, layer.name = "NC Power Plants", 
          zcol = "Plant primary fuel category",
          col.regions = viridis::viridis(length(unique(nc_plants$`Plant primary fuel category`)), 
                                         option = "C")) +
  mapview(nc_counties, layer.name = "NC Counties", col.regions = "grey") 

print(Plant_Location_Color_Map)

#Size by capacity
#Ggplot option
Plant_Location_Size_Plot <- 
  ggplot() +
  geom_sf(data = nc_counties, fill = "grey") +
  geom_sf(data = nc_plants, aes(size = nc_plants$`Plant nameplate capacity (MW)`,
                                color = "red")) +
  labs(title = "Spatial Distribution of Power Plants in NC, Sized by Capacity",
       size = "Capacity (MW)") +
  default_theme +
  theme(legend.position = "bottom")

print(Plant_Location_Size_Plot)

#Mapview option
Plant_Location_Size_Map <- 
  mapview(nc_plants, cex = "Plant nameplate capacity (MW)", 
          zcol = "Plant primary fuel category",
          col.regions = viridis::viridis(length(unique(nc_plants$`Plant primary fuel category`)), 
                                         option = "C"),
          layer.name = 'Plant Primary Fuel Category') +
  mapview(nc_counties, col.regions = "grey", layer.name = 'Counties')

print(Plant_Location_Size_Map)
```

In North Carolina, natural gas power plants are absent from the atlantic coast area as depicted in Figure 3. For the most part, natural gas power plants are instead located slightly inland (though spread across counties).
Solar power plants are evenly throughout the state of North Carolina, depicted, except in Western North Carolina along the Appalachian range. In this mountainous region, hydro power plants are more common than solar power plants.

While the number of solar power plants is large, Figure 4 showcases that each solar plant contains a relatively small electricity generation capacity. The natural gas and coal power plants, though lower in number, producer a larger share of electric power in the state.

```{r EJ Analyses}
#map which counties are classified as disadvantaged

disadvantaged_map <- 
  mapview(EJ_NC_sf, zcol = 'Identified.as.disadvantaged',
        layer.name = 'County Disadvantaged Status')
print(disadvantaged_map)

#Map plants and disadvantaged counties together
disadvantaged_counties_map <- 
  mapview(EJ_NC_sf, zcol = 'Identified.as.disadvantaged',
        layer.name = 'County Disadvantaged Status') +
  mapview(nc_plants, cex = 2,
          layer.name = 'Power Plant Locations')
print(disadvantaged_counties_map)

#Map plants by fuel type and disadvantaged counties
disadvantaged_plant_map <- 
  mapview(EJ_NC_sf, zcol = 'Identified.as.disadvantaged',
        layer.name = 'County Disadvantaged Status') +
  mapview(nc_plants, cex = 4, zcol = "Plant primary fuel category", 
          layer.name = 'Plant Primary Fuel Category') 
print(disadvantaged_plant_map)

```


Figure 5 maps county disadvantage status and power plant location. A few disadvantaged counties have a concentration of power plants in the Southern area of the state. However, the Raleigh metropolitan area is not classified as disadvantaged, but also has a notable concentration of power plants. Therefore, further investigation is required to understand if any correlation exists between county status and power plant locations.

The fuel type of the power plant also impacts front-line communities. Coal and natural gas power plants influence air quality due to the release of criteria air pollutants such as NOx. There are a handful of natural gas and oil power plants located around Robeson and Scotland county, but there are also many solar power plants in these areas. The Raleigh metropolitan area also has a couple of natural gas power plants.

This project aims to explore primary fuel type of electricity generation along with environmental justice to determine if any spatial patterns or injustices exist regarding electricity generation in North Carolina.


\newpage

# Analysis

## Question 1: What is the distribution and capacity of power plants generating electricity in North Carolina?

###Question 1a: Which type of plant has the highest capacity?

The exploratory analysis above investigates which plant fuel category has the highest capacity. In total, gas, solar, nuclear, and coal have the highest generation capacities. However, this does not reflect the average capacity for each fuel type, as shown in figure X below. As displayed in this graph, nuclear has the highest average capacity, followed by coal, gas, wind, hydro, other, biomass, oil, and solar. Although has has the highest total capacity, each individual plant has a lower average capacity than nuclear and coal. This suggests there is a greater quantity of gas plants than nuclear or coal.

###Question 1b: Which type of plant is the most common?
Graph X shows the number of each type of plant. Solar is by far the most common, followed by oil, hydro, biomass, and gas. The reason solar does not have the highest total capacity in MW for NC (despite being the most common) is that it has the lowest average capacity for each plant (Graph X).

###Question 1c: Where are most plants located?
The ten counties with the most power plants are, in order, Robseon, Duplin, Wayne, Nash, Cleveland and Johnston (tied at 29), and Bladen, Catawba, Halifax, and Randolph (all tied at 17). To understand the breakdown of power plants by fuel type in the county with the most plants (Robeson), this county was isolated for analysis. The primary power plant type in Robeson County was solar (41 plants, versus biomass (2) and oil (1)).


```{r}
#Which plant fuel categories have the highest capacity?
print(NC_Plant_Cap_Plot)
#From this plot, it seems gas, solar, nuclear, and coal have the highest capacities.
#Find the average capacity of each fuel type.
average_capacity <- nc_plants %>% 
  group_by(`Plant primary fuel category`) %>% 
  summarize(avg_capacity = mean(`Plant nameplate capacity (MW)`))
#This shows that nuclear has the highest average capacity, followed by coal, gas,
#wind, hydro, other, biomass, oil, and solar. 

#Which plant fuel categories are the most common?
print(FuelType_Count_Plot)
#This plot shows that solar is by far the most common, followed by oil, hydro, 
#biomass, and gas.

#Where are most power plants in NC located?
print(Plant_Location_Color_Map)
#It is difficult to tell from this map.
county_counts <- nc_plants %>% 
  group_by(County.Name) %>% 
  summarize(plant.counts = n())
#Plot the counties with the top 10 highest number of plants.
most_plants_counties <- county_counts %>% 
  filter(plant.counts >= 17)
Plants_by_County_Plot <- ggplot(most_plants_counties, 
                aes(x = fct_reorder(`County.Name`, plant.counts, .desc = TRUE), 
                y = plant.counts, 
                fill = plant.counts)) +
  geom_bar(stat = "identity") +  
  geom_text(aes(label = plant.counts), vjust = -0.5, size = 2.5) +
  labs(title = "Number of Plants per County", x = "County", y = "Number of Plants",
       fill = "County Name") +
  default_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 0.3, size = 8)) +
  scale_fill_viridis(option = "C") 
  
print(Plants_by_County_Plot)
#This plot shows the top 10 counties with the most plants. Robeson county has
#the highest number of plants.

#Plot the breakdown of plant type for Robeson only. 
#First, filter for Robeson.
Robeson <- nc_plants %>% 
  filter(nc_plants$County.Name == "Robeson")
#Find the frequency of each type of fuel category
Robeson_fuel <- Robeson %>%
  group_by(`Plant primary fuel category`) %>%
  summarize(Count = n())

#Now plot by fuel type.
Robeson_Plot <- 
  ggplot(Robeson_fuel, aes(x = `Plant primary fuel category`, y = Count, 
                         fill = `Plant primary fuel category`)) +
  geom_bar(stat = "identity") +  
  geom_text(aes(label = Count), vjust = -0.5, size = 2.5) +
  labs(title = "Number of Plants by Fuel Type", 
       subtitle = "Robeson County",
       x = "Fuel Type", y = "Number of Plants",
       fill = "Plant Primary Fuel Category") +
  default_theme +
  theme(axis.text.x = element_text(angle = 45, hjust = 0.3, size = 8)) +
  scale_fill_viridis_d(option = "C") 

print(Robeson_Plot)
#Robeson County has primarily solar plants.

```

## Question 2: What are the Environmental Justice implications of electricity power generation in North Carolina? 

###Question 2a: Are power plants primarily located in low-income and disadvantaged communities?

```{r}
#Check locations of power plants and disadvantaged counties.
print(disadvantaged_plant_map)
#Hard to tell from this map. Run statistical analyses. 


```

```{r are plant locations significant?}

#join nc plant locations and CEJST data and select relevant columns
EJ.NC_plus_plants <- st_join(EJ_NC_sf, nc_plants_counties, join = st_intersects) %>%
  select(Identified.as.disadvantaged, 
         County.Name, 
         Plant.FIPS.county.code, 
         COUNTYFP, 
         Plant.name,
         Plant.primary.fuel.category,
         Percent.Black.or.African.American.alone,
         Percent.American.Indian...Alaska.Native,
         Percent.Asian,
         Percent.Native.Hawaiian.or.Pacific,
         Percent.two.or.more.races,
         Percent.Hispanic.or.Latino,
         Percent.other.races,
         Percent.White,
         Total.population)

#run a Chi-square test
observed <- table(EJ.NC_plus_plants$Identified.as.disadvantaged)
expected <- table(EJ_NC_sf$Identified.as.disadvantaged)

chisq_test <- chisq.test(observed, p = expected / sum(expected))
chisq_test

#filter nc_plants data set for the type of energy that is the worst for human health 
nc_unhealthy_plants <- EJ.NC_plus_plants %>%
  filter(Plant.primary.fuel.category== 'OIL' | Plant.primary.fuel.category == 'GAS' | Plant.primary.fuel.category== 'COAL' | Plant.primary.fuel.category=='NUCLEAR' |Plant.primary.fuel.category=='BIOMASS')

#Perform Chi-square test for worst types of energy to live by
observed <- table(nc_unhealthy_plants$Identified.as.disadvantaged)
expected <- table(EJ_NC_sf$Identified.as.disadvantaged)

chisq_test <- chisq.test(observed, p = expected / sum(expected))
chisq_test

```


## Question 3: Do Environmental Justice communities face impact from air quality emissions due to power plants?


```{r}
#Examine emissions by fuel type in North Carolina

#view(nc_energy_emissions)

NC_Plant_CO2e_Plot <- ggplot(data = nc_energy_emissions) +
  geom_bar(aes(x = `Plant primary fuel category`,
               y = `total_CO2e`), fill = 'darkred',
           stat = "identity") +
  labs(title = "CO2 Equivalent Emissions from
       Electricity Fuel Type in North Carolina",
       x = "Plant Primary Fuel Category",
       y = "CO2 Equivalent Emissions (Tons)")

print(NC_Plant_CO2e_Plot)

#Create map on CO2 Equivalent Emissions by Fuel Type

mapview(EJ_NC_sf, zcol = 'Identified.as.disadvantaged',
        layer.name = 'County Disadvantaged Status') +
  mapview(nc_plant_emissions,
        cex = "Plant annual CO2 equivalent emissions (tons)",
        zcol = "Plant primary fuel category",
        layer.name = "CO2 Equivalent Emissions (Tons)")

#Check types of other emissions

colnames(nc_plant_emissions)

#Analyze NOx
#EPA defines this as a dangerous pollutant for public health

mapview(EJ_NC_sf, zcol = 'Identified.as.disadvantaged',
        layer.name = 'County Disadvantaged Status') +
  mapview(nc_plant_emissions,
        cex = "Plant annual NOx emissions (tons)",
        zcol = "Plant primary fuel category",
        layer.name = "NOx Emissions (Tons)")

```


\newpage

# Summary and Conclusions


\newpage

# References
<add references here if relevant, otherwise delete this section> 
