---
title: "Electricity Generation and Environmental Justice Communities in North Carolina"
author: "Fiona Price, Summer Heschong, and Asreeta Ushasri"
date: "2024-12-12"
output: html_document
editor_options: 
  chunk_output_type: console
---

\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

```{r setup, include=FALSE}
# Check your working directory

getwd()
here()

# Load your packages

library(dplyr)
library(ggplot2)
library(here)
library(tidyverse)
library(readxl)
library(sf)
library(mapview)

# Set your ggplot theme

default_theme <- theme_light() +
  theme(panel.background = element_rect(color = "white"),
  legend.key = element_rect(color = "white"),
    legend.position = 'right',
    complete = TRUE)

theme_set(default_theme)

# Load your datasets

energy_data <-
  read_csv("./Data/Processed/NC_PowerPlants.csv")

nc_energy_total_cap <-
  read_csv("./Data/Processed/NC_Total_Capacity.csv")

nc_energy_emissions <-
  read_csv("./Data/Processed/NC_Energy_Emissions.csv")

nc_plants <- energy_data %>% 
    st_as_sf(coords = c("Plant longitude","Plant latitude"), 
             crs=4269) 

nc_plant_emissions <- energy_data %>% 
    st_as_sf(coords = c("Plant longitude","Plant latitude"), 
             crs=4269)

nc_EJ_powerplants <- read_csv("./Data/Processed/NC_joined.csv")

EJ.NC_sf <- read_csv("./Data/Processed/NC_Env_Justice.csv") %>%
  st_as_sf()

#County spatial data

nc_counties <- st_read(
  here("./Data/Spatial/North_Carolina_State_and_County_Boundary_Polygons.shp")) 

#Check the CRS
st_crs(nc_counties)$epsg


```


# Rationale and Research Questions

1. What is the distribution and capacity of power plants generating electricity in North Carolina?
2. What are the Environmental Justice implications of electricity power generation in North Carolina? Are power plants primarily located in low-income and disadvantaged communities?
3. Do Environmental Justice communities face impact from carbon dioxide equivalent emissions due to power plants?

\newpage

# Dataset Information

The eGRID 2022 data set provides information on electricity generation in the United States. The U.S. EPA publishes this data to describe capacity and geographic distribution of different energy resources. The 2022 data is the most recent data (published in 2024), because it requires time to process and compile this national data into a workbook.

This project utilizes the Plant 2022 and the Generator 2022 data sheets. The Plant 2022 data sheet describes the capacity, location, and primary fuel source of every power plant in the United States. The Generator 2022 worksheet focuses on individual generators. Power plants often have multiple generators of various fuel types, (such as natural gas combined cycle plants). For this reason, the U.S. EPA includes a separate worksheet for generators specifically.



\newpage

# Exploratory Analysis 



\newpage

# Analysis

## Question 1: What is the distribution and capacity of power plants generating electricity in North Carolina?

```{r}

#view(nc_energy_total_cap)
#colnames(energy_data)

NC_Plant_Cap_Plot <- ggplot(data = nc_energy_total_cap) +
  geom_bar(aes(x = `Plant primary fuel category`,
               y = `total_capacity`), fill = 'darkblue',
           stat = "identity") +
  labs(title = "Power Plant Capacity in North Carolina",
       subtitle = "Asreeta Ushasri",
       x = "Plant Primary Fuel Category",
       y = "Nameplate Capacity (MW)")

print(NC_Plant_Cap_Plot)

```


```{r}
#Mapping capacity of electricity generation in NC

nc_plants$`Plant nameplate capacity (MW)` <-
  as.numeric(nc_plants$`Plant nameplate capacity (MW)`)
class(nc_plants$`Plant nameplate capacity (MW)` )

#view(NC_joined)

#Color by fuel category
ggplot() +
  geom_sf(data = nc_counties, fill = "grey") +
  geom_sf(data = nc_plants, aes(color = nc_plants$`Plant primary fuel category`)) +
  labs(color = "Fuel Category")
  

#Mapview option
mapview(nc_plants, zcol = "Plant primary fuel category") +
  mapview(nc_counties, col.regions = "grey")

```


```{r}
#Size by capacity
#Ggplot option
ggplot() +
  geom_sf(data = nc_counties, fill = "grey", layer.name = 'Counties') +
  geom_sf(data = nc_plants, aes(size = nc_plants$`Plant nameplate capacity (MW)`), layer.name = 'Plant Primary Fuel Category') + 
  theme(legend.position = "none")

#Mapview option
mapview(nc_plants, cex = "Plant nameplate capacity (MW)", zcol = "Plant primary fuel category", layer.name = 'Plant Primary Fuel Category') +
  mapview(nc_counties, col.regions = "grey", layer.name = 'Counties')

```

## Question 2: What are the Environmental Justice implications of electricity power generation in North Carolina? Are power plants primarily located in low-income and disadvantaged communities?

```{r}

#map which counties are classified as disadvantaged

mapview(EJ.NC_sf, zcol = 'Identified.as.disadvantaged',
        layer.name = 'County Disadvantaged Status')

```



```{r}
#Map plants and disadvantaged counties together
mapview(EJ.NC_sf, zcol = 'Identified.as.disadvantaged',
        layer.name = 'County Disadvantaged Status') +
  mapview(nc_plants, cex = 2,
          layer.name = 'Power Plant Locations')

#view(nc_plants)

#Map plants by fuel type and disadvantaged counties
mapview(EJ.NC_sf, zcol = 'Identified.as.disadvantaged',
        layer.name = 'County Disadvantaged Status') +
  mapview(nc_plants, cex = 4, zcol = "Plant.primary.fuel.category", layer.name = 'Plant Primary Fuel Category') 

```


```{r are plant locations significant?}
#run a Chi-square test
observed <- table(EJ.NC_plus_plants$Identified.as.disadvantaged)
expected <- table(EJ.NC_sf$Identified.as.disadvantaged)

chisq_test <- chisq.test(observed, p = expected / sum(expected))
chisq_test

#filter nc_plants data set for the type of energy that is the worst for human health 
nc_unhealthy_plants <- EJ.NC_plus_plants %>%
  filter(Plant.primary.fuel.category== 'OIL' | Plant.primary.fuel.category == 'GAS' | Plant.primary.fuel.category== 'COAL' | Plant.primary.fuel.category=='NUCLEAR' |Plant.primary.fuel.category=='BIOMASS')

#Perform Chi-square test for worst types of energy to live by
observed <- table(nc_unhealthy_plants$Identified.as.disadvantaged)
expected <- table(EJ.NC_sf$Identified.as.disadvantaged)

chisq_test <- chisq.test(observed, p = expected / sum(expected))
chisq_test

```


## Question 3: Do Environmental Justice communities face impact from carbon dioxide equivalent emissions due to power plants?


```{r}
#Examine emissions by fuel type in North Carolina

#view(nc_energy_emissions)

NC_Plant_CO2e_Plot <- ggplot(data = nc_energy_emissions) +
  geom_bar(aes(x = `Plant primary fuel category`,
               y = `total_CO2e`), fill = 'darkred',
           stat = "identity") +
  labs(title = "CO2 Equivalent Emissions from
       Electricity Fuel Type in North Carolina",
       subtitle = "Asreeta Ushasri",
       x = "Plant Primary Fuel Category",
       y = "CO2 Equivalent Emissions (Tons)")

print(NC_Plant_CO2e_Plot)

#Create map on CO2 Equivalent Emissions by Fuel Type

mapview(EJ.NC_sf, zcol = 'Identified.as.disadvantaged',
        layer.name = 'County Disadvantaged Status') +
  mapview(nc_plant_emissions,
        cex = "Plant annual CO2 equivalent emissions (tons)",
        zcol = "Plant primary fuel category",
        layer.name = "CO2 Equivalent Emissions (Tons)")

```


\newpage

# Summary and Conclusions


\newpage

# References
<add references here if relevant, otherwise delete this section> 
